Investigating Neolithic pottery usage from Guijiabao in the Yanyuan Basin in Southwest China using organic residue analysis

### 2023-02-05 analysis/paper/paper.Rmd
added a table for potsherds

---
title: "Investigating Neolithic pottery usage from Guijiabao in the Yanyuan Basin in Southwest China using organic residue analysis"
author:
  - Myfirstname Mylastname:
      email: fl@oneplace.edu
      institute: [UofO]
      correspondence: true
  - Yourfirstname Yourlastname:
      email: fl@another.edu
      institute: [UofA]
      correspondence: false
institute:
  - UofO: University of One Place
  - UofA: University of Another Place
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

@@

The recent growing application of organic residue analysis demonstrates its usefulness to understand ancient pottery use and function [@Kimpe2004; @Mayyas2022], human subsistence strategies, and culinary practices across a region [@Kwak2020; @Craig2011; @Suryanarayan2021; @Zhang2022]. Examining organic residues, particularly lipids, enables archaeologists to investigate ancient food practices through lipid distributions and specific lipid compounds as biomarkers to trace their original foodstuffs [@Copley2005; @Whelton2021]. This is because lipid molecules can be accumulated and well preserved in the porous fabric of pottery over thousand years due to the protection of clay matrices from microbiological degradation and water leaching [@Evershed1993; @Evershed2008; @Roffet2017]. In this study, we detect and identify lipid residues absorbed into the ceramic matrix that can reflect the original contents cooked or stored in a vessel during its use-life [@Dudd1999; @Miller2020]. We attempt to explore the culinary practice under the circumstances during the time of frequent cultural interactions and expansion of agriculture by our case study.

@@

<!--the major finding at GJB-->
The common vessel forms unearthed at Guijiabao include jars, urns, bowls, and basins. For early Neolithic, ceramics have surface decorations, such as incised geometric patterns and impressed dot patterns that resemble ceramic decorations found in western Yunnan, indicative of relationships between Guijiabao and sites in Yunnan [@Guijiabao2017a]. In addition, cord-marked vessels are also found. About the plant remains, the flotation analysis in 2015 reveal crop cultivation dominated by foxtail millet (*Setaria italica*) and broomcorn millet (*Panicum miliaceum*) based on samples collected from ash pits and stratigraphic layers [@Guijiabao2016]. The earliest domesticated millet grain can be traced back to 5000 BP according to its radiocarbon date [@Hao2022]. A smaller number of rice grains (*Oryza sativa*) have also been found in ash pits where the earliest one dates to 4700 BP [@Hao2022]. Later, a detailed phytolith analysis for soil samples dating to the Neolithic period (5000-3700 BP) further suggests the presence of rice phytoliths throughout the temporal sequences, indicative of a mixed millet/rice cultivation prior to 5000 BP [@Huan2022].

@@

## Samples
We selected five ceramic fragments from Guijiabao and one ceramic fragment from Gaoshan for organic residue analysis (Table XX). Four ceramic fragments from Guijiabao were recovered from trench 13 excavated in the 2017 field season and one fragment was from the 2016 season. Those potsherds can be divided into two temporal phases, the early Neolithic period (5000-4500 BP) and the historical period (1000 BP). The common ceramic is cord-marked sand-tempered vessels unearthed at Guijiabao, and their forms include jars, urns, bowls, and basins [@Guijiabao2017a; @Hao2022]. Our samples are mostly reddish-brown jar fragments with medium-size sand temper and some of them have incised cord-marked decoration. Although the original function of those jars is uncertain, they should be supposed to be used for cooking or storage according to the char on the surfaces and their archaeological contexts from a settlement [@Hao2022; @Guijiabao2017a]. This research also attempts to confirm that those ceramics were cooking pots used in the daily life of the local residents.

@@

The GC-MS analysis for organic residues extracted from Guijiabao and Gaoshan potsherds revealed the presence of lipid compounds above interpretable concentrations (5 μg/g) [@Evershed2008experimental], but with low amounts in most of them (Table \@ref(tab:FA-composition-table)). The lipid profiles show that the saturated fatty acids with an even number of carbons, particularly C16:0 (palmitic acid), C18:0 (stearic acid), and monounsaturated fatty acid C18:1 (oleic acid), are the main compounds in most potsherds (Figure \@ref(fig:chromatograms-plots-four)). Those fatty acids commonly exist in animal fat and plant oil. They are also generally degraded from other larger molecules such as triacylglycerols due to oxidation and microbial degradation after long-term burial processes [@Dudd1998; @Eerkens2007]. High concentrations of Medium-chain fatty acids (C16:0 and C18:0) with an even-over-old carbon chain length predominance indicate that the organic residues in our potsherds were highly degraded during post-depositional processes. In addition, a number of potsherds presents higher contents of C18:1 compared to C18:0 and C16:0, such as SYG-TN13-E23-3. Although a larger content of C18:1 and C16:0 with smaller content of C18:0 might indicate an origin of plant oils [@Mayyas2018; @Copley2005], it is hard to directly relate the pattern we observed to a plant origin based on GC-MS results alone.

@@

```{r chromatograms-specific, echo=FALSE, fig.cap= "GC-MS chromatograms of the potshard with the presence of long-chain fatty acids. Cn:x refers to fatty acid with n carbon atoms and x double bonds. IS indicates internal standard used for quantification, including FA C21:0 and n-alkane 34"}
knitr::include_graphics(here::here("analysis/figures/chromatograms_specific.png"))
```

@@

The results show that most potsherd samples from Guijiabao contain fatty acids from ruminant adipose fats with one potsherd (SYG-TN13-E23-3) suggesting an origin of non-ruminant adipose fat (Figure \@ref(fig:isotopic-plots): A). The potsherd from Gaushan (CDG-062) demonstrates ruminant adipose fats as well. A number of potsherds (SYG-TN13-E22-1 and SYG-3) from Guijiabao dating to the historical period plotted between the reference line of ruminant adipose and non-ruminant, indicative of the mixing of those two origins [@Copley2003; @Dudd1998milk; @Suryanarayan2021]. Moreover, the potsherd indicating non-ruminant adipose fats has a relatively depleted δ^13^C value (Figure \@ref(fig:isotopic-plots): B) falling within the range of C3/terrestrial plants or freshwater fish [@Shoda2018; @Kwak2020]. This also corresponds to the lipid distribution of the same potshard which reflects a possible plant origin based on the GC-MS results. Although the referenced isotopic values should be ideally adjusted in terms of local environments, they could still provide useful information for accessing food categories.

@@

# Discussion
<!--A brief summary of interesting results-->
Our results show that food choices indicated by lipid residues extracted from Guijiabao and Gaoshan potsherds were mainly dominated by ruminant fats during the Neolithic period. Moreover, it is striking to identify a non-ruminant source from a Guijiabao potshard that might originate from C3 plants. Although the lipid yields in most of our samples are generally low, interpretative amounts of yields from all of our samples support that those ceramics were used as cooking or storage pots. The low yield might be related to the part of pottery samples analyzed in this study are mostly base fragments [@Evershed2008experimental; @Kwak2020]. But we also found a base sample dates to early Neolithic period contains a rich composition of long-chain fatty acids, reflecting a good preservation. We recognize the limitation for interpreting food choices over time or differences across regions because of our small sample sizes. Also, the lack of local modern references for animal fats and plants has impacts on our explanation. However, the general pattern of the isotopic values of lipid molecules could still provide informative clues for a better understanding of culinary practices or the local subsistence at Guijiabao.

<!--discussion about possible plant origins-->
One aim of our study is to explore whether new crops, such as C3 plants (eg. rice), C4 plants (eg. millets), or new species (eg. pigs/horses) became an essential part of local diet in Yanyuan during the Neolithic transition. We successfully identified non-ruminant fats from a Guijiabao potshard that might be from terrestrial plant/C3 plants based on similar isotopic values reported by other studies in East Asia [@Shoda2018; @Kwak2020]. It is also possible other native grasses classified as C3 plants, such as pennisetum and chenopodiaceae, were part of the local diet [@Guijiabao2016]. However, the seeds of wild weeds only account for a small part of plant remains. Surprisingly, we only observed signals of C3 plants, despite the fact that C4 plants (foxtail and broomcorn millets) are the major crop based on the recent macro-botanical and phytolith studies at Guijiabao [@Guijiabao2016; @Hao2022; @Huan2022]. Although freshwater fish could be another origin of non-ruminant fats that can suggested by the finding of fish bones and net sinkers at Guijiabao [@Lucquin2016; @Guijiabao2017b], we did not detect any presence of biomarkers for aquatic resources, such as long-chain $\omega$-(o-alkylphenyl) alkanoic acids or phytanic acid from our potsherds [@Evershed2008marine; @Hansel2004; @Cramp2014; @Kwak2017].

<!--discussion about possible ruminant fat origin-->
Ruminant fats identified in the Neolithic potsherds (ca. 4500 BP) from Guijiabao and Gaoshan could be derived from deer or horse according to unearthed faunal remains from the sites, accompanied with stone tools for hunting, such as stone balls and arrow-heads [@Hao2022; @Guijiabao2020a]. But whether horse was part of local food choices in the Neolithic period is questionable. First, the Pleistocene and Neolithic sites with wild horse bones primarily concentrated in northern China [@Yuan2006]. Second, domesticated horses first found in southwest China were much later from the Dian culture (4th century BC to 1st century AD) in eastern Yunnan [@Lei2009; @Chiou2004]. The evidence from Guijuabao also confirms a later presence of horse remains. In addition, one potsherd (SYG-TN13-E22-2#1) shows the isotopic value consistent with those values of deer and sheep fats fed on C3 diet based on global references [@Evershed2002]. It is also possible that other ruminant animals, such as wild Asian water buffalo or yak, were part of food sources during this period [@Zhang2016]. A detailed analysis of faunal remains from Guijiabao is still in progress that can provide further information.

We observed a mixture of ruminant fats and terrestrial plants based on a number of Guijiabao potsherds dating to the historical period (ca. 1000 BP). However, we are unable to determine whether there is a shift in diet over time due to our limited sample size. But the overall depleted δ^13^C values for most potsherds suggest C3 plants might be an important diet for local animals and could also be included in the local diet in Yanyuan during the Neolithic transition. This corresponds with the timing when new crops spread into southwest China. Despite of no C4 plant origin in our samples, one potshard implies a source of C4 plant-fed ruminant. In general, the food source detected from our samples comprised mainly of ruminant fats that are more likely from wildlife. This might be related to local ecological zones and environments where wild animals were easier to obtain. Since zooarchaeological analysis and other studies on stone tools are still in progress, we could only rely on preliminary excavation reports with limited details about associated artifacts. A large-scale organic residue analysis of ceramics across different pottery forms in a wider region may offer insight into temporal and regional variations in food choices during the Neolithic transition.

@@

Our case study on organic residues preserved in pottery fragments from Guijiabao and Gaoshan in southwest China reveals the presence of lipids but in low concentrations for most potsherds. The food cooked in those ceramics include ruminant fats and terrestrial plants during the Neolithic transition when the mixing farming of rice/millet spread into southwest China. It is most likely the terrestrial plant signal is C3 plant, implying rice processing in the pot. However, the evidence of direct processing of terrestrial plants is not as much as we expected according to a relatively large quantity of plant remains found at Guijiabao, such as millet grains. Instead, food processing in the early Neolithic is mostly ruminant fats that could originate from local wild animals, such as deer. This indicates that the proportion of rice/millet in the local diet is not high during the Neolithic transition, suggesting that hunting might be the primary subsistence, supplemented by rice/millet farming. We recognize the limitation of our interpretation of the local diet due to the small sample size. However, our study can still offer a perspective from organic residue analysis to explore food preferences and the extent of farming adoption when rice/millet agriculture first spread into the Yanyuan during the Neolithic period. A more systematic sampling of potsherds across different phases and pottery types may eliminate a clear picture of diet and subsistence during the Neolithic transition.

>>>>

---
title: "Investigating Neolithic pottery usage from Guijiabao in the Yanyuan Basin in Southwest China using organic residue analysis"
author:
  - Li-Ying Wang Mylastname:
      email: fl@oneplace.edu
      institute: [UofO]
      correspondence: true
  - Yourlastname:
      email: fl@another.edu
      institute: [UofA]
      correspondence: false
institute:
  - UofO: University of One Place
  - UofA: University of Another Place
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

@@

The recent growing application of organic residue analysis demonstrates its usefulness to understand ancient pottery use and function [@Kimpe2004; @Mayyas2022], human subsistence strategies, and culinary practices across a region [@Kwak2020; @Craig2011; @Suryanarayan2021; @Zhang2022]. Examining organic residues, particularly lipids, enables archaeologists to investigate ancient food practices through lipid distributions and specific lipid compounds as biomarkers to trace their original foodstuffs [@Copley2005; @Whelton2021]. This is because lipid molecules can be accumulated and well preserved in the porous fabric of pottery over thousand years due to the protection of clay matrices from microbiological degradation and water leaching [@Evershed1993; @Evershed2008; @Roffet2017]. In this study, we detect and identify lipid residues absorbed into the ceramic matrix that can reflect the original contents cooked or stored in a vessel during its use-life [@Dudd1999; @Miller2020]. We attempt to explore the culinary practice under the circumstances during the time of frequent cultural interactions and expansion of agriculture through our case study.

@@

<!--the major finding at GJB-->
The common vessel forms unearthed at Guijiabao include jars, urns, bowls, and basins. The common ceramic styles in the early Neolithic period include grayish high-tempered fine ware and reddish brown coarse ware with surface decorations such as cord-marked and basket-impressed patterns [@Hao2022; @Guijiabao2017a]. Among the ceramics, incised geometric patterns and impressed dot patterns that resemble ceramic decorations found in western Yunnan, indicative of relationships between Guijiabao and sites in Yunnan [@Guijiabao2017a]. About the plant remains, the flotation analysis in 2015 reveal crop cultivation dominated by foxtail millet (*Setaria italica*) and broomcorn millet (*Panicum miliaceum*) based on samples collected from ash pits and stratigraphic layers [@Guijiabao2016]. The earliest domesticated millet grain can be traced back to 5000 BP according to its radiocarbon date [@Hao2022]. A smaller number of rice grains (*Oryza sativa*) have also been found in ash pits where the earliest one dates to 4700 BP [@Hao2022]. Later, a detailed phytolith analysis for soil samples dating to the Neolithic period (5000-3700 BP) further suggests the presence of rice phytoliths throughout the temporal sequences, indicative of a mixed millet/rice cultivation prior to 5000 BP [@Huan2022].

@@

## Samples
We selected six ceramic fragments from Guijiabao and Gaoshan for organic residue analysis (Table \@ref(tab:sherds-table)). Four ceramic fragments from Guijiabao were recovered from trench 13 excavated in the 2017 field season and one fragment was from the 2016 season. Those potsherds can be divided into two temporal phases, the early Neolithic period (ca. 4500 BP) and the historical period (ca. 1000 BP). Our samples are mostly grayish base fragments with fine sand temper and a number of them have impressed decoration. Although their original forms could not be recovered, those ceramics are supposed to be used in daily life according to the evidence of chars on some fragments and their archaeological contexts in cultural layers [@Hao2022; @Guijiabao2017a]. This research also attempts to confirm that those ceramics were used as cooking/storage wares for food.

```{r sherds-table, cache= TRUE, cache.extra=tools::md5sum(here::here("analysis/data/raw_data/SW-china-potsherd.xlsx"))}
Ch_sherds <-
  readxl::read_excel(here::here("analysis", "data", "raw_data", "SW-china-potsherd.xlsx"), col_names = TRUE)

knitr::kable(Ch_sherds,
             caption = "Quantitative data of fatty acid (FA) composition (ug) of potsherds from Guijiabao and Gaushan that are named starting with 'SYG' and 'CDG' respectively")
```

@@

The GC-MS analysis for organic residues extracted from Guijiabao and Gaoshan potsherds revealed the presence of lipid compounds above interpretable concentrations (5 μg/g) [@Evershed2008experimental], but with low amounts in most of them (Table \@ref(tab:FA-composition-table)). The lipid profiles show that the saturated fatty acids with an even number of carbons, particularly C16:0 (palmitic acid), C18:0 (stearic acid), and monounsaturated fatty acid C18:1 (oleic acid), are the main compounds in most potsherds (Figure \@ref(fig:chromatograms-plots-four)). Those fatty acids commonly exist in animal fat and plant oil. They are also generally degraded from other larger molecules such as triacylglycerols due to oxidation and microbial degradation after long-term burial processes [@Dudd1998; @Eerkens2007]. High concentrations of Medium-chain fatty acids (C16:0 and C18:0) with an even-over-old carbon chain length predominance indicate that the organic residues in our potsherds were highly degraded during post-depositional processes. In addition, a number of potsherds present higher contents of C18:1 compared to C18:0 and C16:0, such as SYG-TN13-E23-3. Although a larger content of C18:1 and C16:0 with smaller content of C18:0 might indicate an origin of plant oils [@Mayyas2018; @Copley2005], it is hard to directly relate the pattern we observed to a plant origin based on GC-MS results alone.

@@

```{r chromatograms-specific, echo=FALSE, fig.cap= "GC-MS chromatograms of the potsherd with the presence of long-chain fatty acids. Cn:x refers to fatty acid with n carbon atoms and x double bonds. IS indicates internal standard used for quantification, including FA C21:0 and n-alkane 34"}
knitr::include_graphics(here::here("analysis/figures/chromatograms_specific.png"))
```

@@

The results show that most potsherd samples from Guijiabao contain fatty acids from ruminant adipose fats with one potsherd (SYG-TN13-E23-3) suggesting an origin of non-ruminant adipose fat (Figure \@ref(fig:isotopic-plots): A). The potsherd from Gaushan (CDG-062) demonstrates ruminant adipose fats as well. A number of potsherds (SYG-TN13-E22-1 and SYG-3) from Guijiabao dating to the historical period plotted between the reference line of ruminant adipose and non-ruminant, indicative of the mixing of those two origins [@Copley2003; @Dudd1998milk; @Suryanarayan2021]. Moreover, the potsherd indicating non-ruminant adipose fats has a relatively depleted δ^13^C value (Figure \@ref(fig:isotopic-plots): B) falling within the range of C3/terrestrial plants or freshwater fish [@Shoda2018; @Kwak2020]. This also corresponds to the lipid distribution of the same potsherd which reflects a possible plant origin based on the GC-MS results. Although the referenced isotopic values should be ideally adjusted in terms of local environments, they could still provide useful information for accessing food categories.

@@

# Discussion
<!--A brief summary of interesting results-->
Our results show that food choices indicated by lipid residues extracted from Guijiabao and Gaoshan potsherds were mainly dominated by ruminant fats during the Neolithic period. Moreover, it is striking to identify a non-ruminant source from a Guijiabao potsherd that might originate from C3 plants. Although the lipid yields in most of our samples are generally low, interpretative amounts of yields from all of our samples support that those ceramics were used as cooking or storage pots. The low yield might be related to the part of pottery samples analyzed in this study are mostly base fragments [@Evershed2008experimental; @Kwak2020]. But we also found a base sample dating to the early Neolithic period contains a rich composition of long-chain fatty acids, reflecting good preservation. We recognize the limitation for interpreting food choices over time or differences across regions because of our small sample sizes, and a lack of local modern references for animal fats and plants. However, the general pattern of the isotopic values of lipid molecules could still provide informative clues for a better understanding of culinary practices or the local subsistence at Guijiabao.

<!--discussion about possible plant origins-->
One aim of our study is to explore whether new crops, such as C3 plants (eg. rice), C4 plants (eg. millets), or new species (eg. pigs/horses) became an essential part of the local diet in Yanyuan during the Neolithic transition. We successfully identified non-ruminant fats from a Guijiabao potsherd that might be from terrestrial plant/C3 plants based on similar isotopic values reported by other studies in East Asia [@Shoda2018; @Kwak2020]. It is also possible other native grasses classified as C3 plants, such as pennisetum and chenopodiaceae, were part of the local diet [@Guijiabao2016]. However, the seeds of wild weeds only account for a small part of plant remains. Surprisingly, we only observed signals of C3 plants, despite the fact that C4 plants (foxtail and broomcorn millets) are the major crop based on the recent macro-botanical and phytolith studies at Guijiabao [@Guijiabao2016; @Hao2022; @Huan2022]. Freshwater fish could be another origin of non-ruminant fats based on the finding of fish bones and net sinkers at Guijiabao [@Lucquin2016; @Guijiabao2017b], but we did not detect any presence of biomarkers for aquatic resources, such as long-chain $\omega$-(o-alkylphenyl) alkanoic acids or phytanic acid from our potsherds [@Evershed2008marine; @Hansel2004; @Cramp2014; @Kwak2017].

<!--discussion about possible ruminant fat origin-->
Ruminant fats identified in the Neolithic potsherds (ca. 4500 BP) from Guijiabao and Gaoshan could be derived from deer or horse according to unearthed faunal remains from the sites, accompanied with stone tools for hunting, such as stone balls and arrow-heads [@Hao2022; @Guijiabao2020a]. But whether horse was part of local food choices in the Neolithic period is questionable. First, the Pleistocene and Neolithic sites with wild horse bones primarily concentrated in northern China [@Yuan2006]. Second, domesticated horses first found in southwest China were much later from the Dian culture (4th century BC to 1st century AD) in eastern Yunnan [@Lei2009; @Chiou2004]. The evidence from Guijuabao also confirms a later presence of horse remains. In addition, one potsherd (SYG-TN13-E22-2#1) shows the isotopic value consistent with those values of deer and sheep fats fed on a C3 diet based on global references [@Evershed2002]. It is also possible that other ruminant animals, such as wild Asian water buffalo or yak, were part of food sources during this period [@Zhang2016]. A detailed analysis of faunal remains from Guijiabao is still in progress that can provide further information.

We observed a mixture of ruminant fats and terrestrial plants based on a number of Guijiabao potsherds dating to the historical period (ca. 1000 BP). However, we are unable to determine whether there is a shift in diet over time due to our limited sample size. But the overall depleted δ^13^C values for most potsherds suggest C3 plants might be an important diet for local animals and could also be included in the local diet in Yanyuan during the Neolithic transition. This corresponds with the timing when new crops spread into southwest China. Despite of no C4 plant origin in our samples, one potsherd implies a source of C4 plant-fed ruminant. In general, the food source detected from our samples comprised mainly of ruminant fats that are more likely from wildlife. This might be related to local ecological zones and environments where wild animals were easier to obtain. Since zooarchaeological analysis and other studies on stone tools are still in progress, we could only rely on preliminary excavation reports with limited details about associated artifacts. A large-scale organic residue analysis of ceramics across different pottery forms in a wider region may offer insight into temporal and regional variations in food choices during the Neolithic transition.

@@

Our case study on organic residues preserved in pottery fragments from Guijiabao and Gaoshan in southwest China reveals the presence of lipids but in low concentrations for most potsherds. The food cooked in those ceramics include ruminant fats and terrestrial plants during the Neolithic transition when the mixing farming of rice/millet spread into southwest China. It is most likely the terrestrial plant signal from C3 plant, implying rice processing in the pot. However, the evidence of direct processing of terrestrial plants is not as much as we expected according to a relatively large quantity of plant remains found at Guijiabao, such as millet grains. Instead, food processing in the early Neolithic is mostly ruminant fats that could originate from local wild animals, such as deer. This indicates that the proportion of rice/millet in the local diet is not high during the Neolithic transition, suggesting that hunting might be the primary subsistence, supplemented by rice/millet farming. We recognize the limitation of our interpretation of the local diet due to the small sample size. However, our study can still offer a perspective from organic residue analysis to explore food preferences and the extent of farming adoption when rice/millet agriculture first spread into the Yanyuan during the Neolithic period. A more systematic sampling of potsherds across different phases and pottery types may eliminate a clear picture of diet and subsistence during the Neolithic transition.
